A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-30 paint-contour-tree.tex
changes to intro

\newcommand{\bT}{{\bf T}}

@@

Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
A classic result of Carr, Snoeyink, and Axen (2000) give an algorithm that takes $O(n\log n + N\alpha(N))$ time
(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N\alpha(N))$ time
was given by Chiang et al. All these algorithms involve a global sort of the critical points,
which is a major bottleneck for large datasets. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + N\alpha(N))$. If the contour tree is a balanced
binary tree, our running time is $O(n + N\alpha(N))$. Moreover, we prove a lower bound of
exactly $\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, thereby showing
that the tree structure must play a role in the time complexity.

@@

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold 
(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.
}

In order to produce our instance optimal results we introduce a novel linear time surface partitioning scheme which is likely to have applications elsewhere.
Moreover, in order to prove the optimality we use a non-trivial combinatorial discharging argument which we believe is of independent interest and in particular uses 
such tools as heavy-light path decompositions.
\end{abstract}

@@

\begin{theorem} \label{thm:alg-main} Consider $f:\MM \mapsto \RR$ that is PL-Morse, where the contour tree $T$
has maximum degree\footnote{The maximum degree $3$ assumption is primarily there to simplify the presentation and proof. Our algorithm
does not need the assumption.} $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$ and any $p \in P(T)$, obviously $|p| \leq |T| = t$.
So the running time is clearly $O(t\log t + N)$. But this is only the case when $T$ is tall and thin.  
An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{alg-main}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

>>>>

\newcommand{\bF}{{\bf F}}
\newcommand{\bT}{{\bf T}}

@@

Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
A classic result of Carr, Snoeyink, and Axen (2000) give an algorithm that takes $O(n\log n + N\alpha(N))$ time
(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
was given by Chiang et al. All these algorithms involve a global sort of the critical points,
which is a major bottleneck for large datasets. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t))$. Moreover, we prove a general lower bound of
$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
that the tree structure must play a role in the time complexity.

@@

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold 
(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.

In order to produce our instance optimal results we introduce a novel linear time surface partitioning scheme which is likely to have applications elsewhere.
Moreover, in order to prove the optimality we use a non-trivial combinatorial discharging argument which we believe is of independent interest and in particular uses 
such tools as heavy-light path decompositions.
}
\end{abstract}

@@

\begin{theorem} \label{thm:main-alg} Consider $f:\MM \mapsto \RR$ that is PL-Morse, where the contour tree $T$
has maximum degree\footnote{The maximum degree $3$ assumption is primarily there to simplify the presentation and proof. Our algorithm
does not need the assumption.} $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$ and any $p \in P(T)$, obviously $|p| \leq |T| = t$.
So the running time is clearly $O(t\log t + N)$. But this is only the case when $T$ is tall and thin.  
An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

We complement \Thm{main-alg} with an almost matching lower bound. We show that the complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case. For any $f$,
we use $\cost(f)$ to denote $\sum_{p \in P(T)} |p|\log |p|$ for the specific $P(T)$
given in \Thm{main-alg}. Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for $f$. But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
a refined lower bound matching \Thm{main-alg}. A \emph{height decision tree} is an algorithm
that is allowed aribtrary computation on the domain $\MM$, and is an algebraic decision
tree over the heights. The running time is simply the number of queries made.

\begin{theorem} \label{thm:main-lb} For any valid cost $C$, there exists a set $\bF$ of functions over 2-dimensions
with the following properties.
For all $f \in \bF$, $\cost(f) = C$. Any height decision tree that correctly computes
the contour tree on all of $\bF$ has an average case (uniform over $\bF$) running time of $\Omega(C)$.
\end{theorem}

In other words, for any valid cost $C$, we can find a family of inputs where \Thm{main-alg}
guarantees a running time of $C + t\alpha(t)$, and any height decision tree requires $\Omega(C)$ time.
This shows that the cost function is inherent to the computation of contour trees.
